Concurrent Programming in Harmony

### 2020-09-26 book/paper.tex
Latest

However, if you run the program through Harmony (\autoref{fig:naiveflagshtml}),
it turns out the solution
does has a problem: if both try to enter the critical section at the same
time, they may end up waiting for one another indefinitely.  Thus the
solution violates \emph{progress}.

@@

So we are looking for an inductive invariant that \emph{implies} exclusivity.
In other words, the set of states where the inducive invariant holds
must be a subset of the set of states where there is at most one process in
the critical section.
Let $\mathcal{C}(i) = \lnot\mathit{flags}[1 - i] \lor
\mathit{turn} = i$, that is, the condition on the \textbf{await}
statement for process~$i$.
Let predicate $\mathcal{I}_p(i)$ be the following:
if process $i$ is at label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $\mathcal{C}(i)$ holds or process $1-i$ is executing after setting
$\mathit{flags}[i]$ but still before setting \textit{turn} to $1-i$.
More formally, $\mathcal{I}_p(i) = \mathtt{process}(i)@cs \Rightarrow \mathcal{C}(i) \lor \mathtt{process}(1-i)@gate$.
For Peterson's Algorithm, an inductive invariant that works well is
$\mathcal{I}_p(0) \land \mathcal{I}_p(1)$.

@@

\begin{figure}
\begin{center}
\includegraphics[width=2.3in]{figures/rdwr-crop.pdf}
\end{center}
\caption{High-level state diagram specification of reader/writer locks with
up to two processes.
The first number in a state gives the number of processes; the second number is the
number of processes reading in the critical section; the third is the number of
processes writing in the critical section.}
\label{fig:rdwr}
\end{figure}

@@

\section*{Exercises}
\begin{problems}
\item Draw additional states and steps in \autoref{fig:rdwr}
for three processes.
\item Optimize your solutions to \autoref{ex:qcontains} to use reader/writer locks.
\item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
up to \texttt{M} processes to acquire it at the same time.  A bound lock
with \texttt{M == 1} is an ordinary lock.
You should define a constant \texttt{M} and two methods:
\texttt{acquire\_bound\_lock()}
and \texttt{release\_bound\_lock()}.
(Bound locks are useful for situations where too many processes working
at the same time might exhaust some resource such as a cache.)
\item Write a test program for your bound lock
that checks that no more than \texttt{M} processes can acquire the
bound lock.
\end{problems}

@@

\begin{itemize}
\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathit{nreaders} \ge n$,
\item $\exists$ writer at \texttt{@wcs} $\Rightarrow \mathit{nreaders} = 0$,
\end{itemize}

A writer simply acquires \textit{rwlock} to enter the critical section
and releases it to exit.  The \emph{first} reader to enter the critical
section acquires \textit{rwlock} and the \emph{last} reader to exit
the critical section releases \textit{rwlock}.
The implementation satisfies progress: if no process is in the critical
section then any process can enter.

>>>>

However, if you run the program through Harmony (\autoref{fig:naiveflagshtml}),
it turns out the solution
does have a problem: if both try to enter the critical section at the same
time, they may end up waiting for one another indefinitely.  Thus the
solution violates \emph{progress}.

@@

So we are looking for an inductive invariant that \emph{implies} exclusivity.
In other words, the set of states where the inductive invariant holds
must be a subset of the set of states where there is at most one process in
the critical section.
Let $\mathcal{C}(i) = \lnot\mathit{flags}[1 - i] \lor
\mathit{turn} = i$, that is, the condition on the \textbf{await}
statement for process~$i$.
Let predicate $\mathcal{I}_p(i)$ be the following:
if process $i$ is at label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $\mathcal{C}(i)$ holds or process $1-i$ is executing after setting
$\mathit{flags}[i]$ but still before setting \textit{turn} to $1-i$.
More formally, $\mathcal{I}_p(i) = \mathtt{process}(i)@cs \Rightarrow \mathcal{C}(i) \lor \mathtt{process}(1-i)@gate$.
For Peterson's Algorithm, an inductive invariant that works well is
$\mathcal{I}_p(0) \land \mathcal{I}_p(1)$.

@@

\begin{figure}
\begin{center}
\includegraphics[width=2.3in]{figures/rdwr.png}
\end{center}
\caption{High-level state diagram specification of reader/writer locks with
up to two processes.
The first number in a state gives the number of processes; the second number is the
number of processes reading in the critical section; the third is the number of
processes writing in the critical section.}
\label{fig:rdwr}
\end{figure}

@@

\section*{Exercises}
\begin{problems}
\item Draw additional states and steps in \autoref{fig:rdwr}
for three processes.
\item Optimize your solutions to \autoref{ex:qcontains} to use reader/writer locks.
\item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
up to \texttt{M} processes to acquire it at the same time.\footnote{A bound lock is a restricted version of a counting semaphore, as we will learn later in this book.}
A bound lock
with \texttt{M == 1} is an ordinary lock.
You should define a constant \texttt{M} and two methods:
\texttt{acquire\_bound\_lock()}
and \texttt{release\_bound\_lock()}.
(Bound locks are useful for situations where too many processes working
at the same time might exhaust some resource such as a cache.)
\item Write a test program for your bound lock
that checks that no more than \texttt{M} processes can acquire the
bound lock.
\end{problems}

@@

\begin{itemize}
\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathit{nreaders} \ge n$
\item $\exists$ writer at \texttt{@wcs} $\Rightarrow \mathit{nreaders} = 0$
\end{itemize}

A writer just acquires \textit{rwlock} to enter the critical section
and releases it to exit.  The \emph{first} reader to enter the critical
section acquires \textit{rwlock} and the \emph{last} reader to exit
the critical section releases \textit{rwlock}.
The implementation satisfies progress: if no process is in the critical
section then any process can enter.
